---
title: "BOSTON 311 SERVICE REQUEST DATA"
subtitle: "Mid-Term 1"
author: "Moses Ayirebi"
date: "February 4, 2025"
output: 
  pdf_document:
    latex_engine: xelatex
geometry: margin=0.9in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(lubridate)
library(scales)
```

```{r load-data}
boston_311 <- read_csv("boston_311.csv.csv")

boston_311 <- boston_311 %>%
  mutate(
    response_time_days = as.numeric(CLOSED_DT - OPEN_DT),
    day_of_week = wday(OPEN_DT, label = TRUE, abbr = FALSE),
    case_status = case_when(
      response_time_days < 0 ~ "Impossible",
      response_time_days > 365 ~ "Zombie",
      is.na(response_time_days) ~ "Still Open",
      TRUE ~ "Valid"
    )
  )
```

## Overview

When Boston residents report a pothole, request graffiti removal, needle removal or flag an abandoned vehicle, how quickly does the city respond and does it depend on where you live? This Read-Me Document explores whether response times reveal patterns of service delivery across neighborhoods.

This document introduces the Boston 311 Service Request Dataset from the City of Boston's Analyze Boston data portal. The dataset contains `r format(nrow(boston_311), big.mark = ",")` service requests submitted by residents between 2020 and 2024. Each row represents one request for city services and includes `r ncol(boston_311)` variables capturing request type, timing, location, and resolution status.

The analysis uses R functions including `median()`, `mean()`, `max()`, `filter()`, `n_distinct()`, `count()`, `table()`, and `wday()` for data exploration, and `geom_col()` and `geom_histogram()` for visualization.

## Fun Facts

**1.** Half of all requests close within `r median(boston_311$response_time_days, na.rm = TRUE)` day, but the average is `r round(mean(boston_311$response_time_days, na.rm = TRUE), 1)` days, meaning a small number of extremely delayed cases drag up the mean dramatically. *Functions: `median()`, `mean()`*

**2.** South Dorchester generates the most 311 requests (`r format(boston_311 %>% filter(BRA_PD == "South Dorchester") %>% nrow(), big.mark = ",")` total, `r round(boston_311 %>% filter(BRA_PD == "South Dorchester") %>% nrow() / nrow(boston_311) * 100, 1)`% of all cases)- but is this because the neighborhood has more problems, or because residents there are more likely to report issues? *Functions: `filter()`, `count()`*

**3.** The 311 system uses `r n_distinct(boston_311$REASON)` broad request categories. The most common is "`r boston_311 %>% count(REASON, sort = TRUE) %>% slice(1) %>% pull(REASON)`." *Functions: `n_distinct()`, `count()`*

**4.** Weekday requests account for `r round(boston_311 %>% filter(!day_of_week %in% c("Saturday", "Sunday")) %>% nrow() / nrow(boston_311) * 100, 1)`% of all submissions. Tuesday is busiest (`r round(boston_311 %>% filter(day_of_week == "Tuesday") %>% nrow() / nrow(boston_311) * 100, 1)`%), Sunday slowest (`r round(boston_311 %>% filter(day_of_week == "Sunday") %>% nrow() / nrow(boston_311) * 100, 1)`%). *Functions: `wday()`, `count()`*

**5.** The longest time to close a request was `r format(max(boston_311$response_time_days, na.rm = TRUE), big.mark = ",")` days—over `r round(max(boston_311$response_time_days, na.rm = TRUE) / 365, 1)` years. *Function: `max()`*

**6.** There are `r format(boston_311 %>% filter(response_time_days < 0) %>% nrow(), big.mark = ",")` records with negative response times, a data quality issue analysts should filter out. *Functions: `filter()`, `nrow()`*

**7.** Over `r format(boston_311 %>% filter(response_time_days > 365) %>% nrow(), big.mark = ",")` requests took more than one year to close, suggesting systemic delays. *Functions: `filter()`, `nrow()`*

**8.** The top three request sources are `r boston_311 %>% count(SOURCE, sort = TRUE) %>% slice(1) %>% pull(SOURCE)`, `r boston_311 %>% count(SOURCE, sort = TRUE) %>% slice(2) %>% pull(SOURCE)`, and `r boston_311 %>% count(SOURCE, sort = TRUE) %>% slice(3) %>% pull(SOURCE)`. *Function: `table()`*

\newpage

## Visualizations

### Figure 1: Mean Response Time by Neighborhood

This chart compares average response times across Boston's ten highest-volume neighborhoods. Using the mean (instead of median) reveals variation hidden by the median's 0-1 day range. *Function: `geom_col()`*

```{r viz-1, fig.width=6, fig.height=3.8, fig.align='center'}
top_neighborhoods <- boston_311 %>%
  filter(!is.na(BRA_PD)) %>%
  count(BRA_PD, sort = TRUE) %>%
  head(10) %>%
  pull(BRA_PD)

neighborhood_stats <- boston_311 %>%
  filter(BRA_PD %in% top_neighborhoods, 
         response_time_days >= 0, 
         response_time_days <= 365) %>%
  group_by(BRA_PD) %>%
  summarise(
    mean_response = mean(response_time_days, na.rm = TRUE),
    request_count = n()
  )

neighborhood_stats %>%
  ggplot(aes(x = reorder(BRA_PD, mean_response), y = mean_response)) +
  geom_col(aes(fill = mean_response), color = "white", width = 0.7) +
  geom_text(aes(label = paste0(round(mean_response, 1), " days")), 
            hjust = -0.1, size = 2.5, fontface = "bold") +
  geom_text(aes(label = paste0("n=", format(request_count, big.mark = ","))), 
            y = 0.02, hjust = 0, size = 2, color = "white") +
  coord_flip() +
  scale_fill_gradient(low = "#9B59B6", high = "#E74C3C", guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Average Response Times Vary Across Neighborhoods",
       subtitle = "Top 10 neighborhoods by volume",
       x = NULL, y = "Mean Response Time (Days)") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 8, color = "gray40"),
        axis.text.y = element_text(size = 8, face = "bold"),
        panel.grid.major.y = element_blank())
```
Neighborhoods with higher mean response times, like East Boston and Jamaica Plain, likely have more "zombie cases" dragging up their averages. This raises an important equity question: are some communities waiting longer for city services, or do certain request types concentrated in those areas simply take longer to resolve?

 \newpage 
 
### Figure 2: Distribution of Response Times

Most requests are resolved quickly, but a long tail reveals some take weeks. *Function: `geom_histogram()`*

```{r viz-2, fig.width=6, fig.height=3.2, fig.align='center'}
valid_responses <- boston_311$response_time_days[
  boston_311$response_time_days >= 0 & 
  boston_311$response_time_days <= 30]

median_val <- median(valid_responses, na.rm = TRUE)
mean_val <- mean(valid_responses, na.rm = TRUE)

boston_311 %>%
  filter(response_time_days >= 0, response_time_days <= 30) %>%
  ggplot(aes(x = response_time_days)) +
  geom_histogram(binwidth = 1, fill = "#9B59B6", color = "white") +
  geom_vline(xintercept = median_val, linetype = "dashed", 
             color = "#E74C3C", linewidth = 1) +
  geom_vline(xintercept = mean_val, linetype = "dotted", 
             color = "#3498DB", linewidth = 1) +
  annotate("label", x = 15, y = 500000, 
           label = paste0("Median = ", round(median_val, 1), " day"), 
           fill = "#E74C3C", color = "white", 
           fontface = "bold", size = 2.5, label.size = 0) +
  annotate("label", x = 15, y = 380000, 
           label = paste0("Mean = ", round(mean_val, 1), " days"), 
           fill = "#3498DB", color = "white", 
           fontface = "bold", size = 2.5, label.size = 0) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Most Requests Resolved Within One Week",
       subtitle = "Distribution of response times (0-30 days)",
       x = "Response Time (Days)", y = "Number of Requests") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 8, color = "gray40"))
```

The sharp spike at 0-1 days shows Boston's 311 system resolves most requests almost immediately—likely automated closures or quick fixes. However, the gap between the median (1 day) and mean (2.4 days) tells a deeper story: while half of all cases close within a day, the long tail of slower cases pulls the average higher. This histogram only shows requests up to 30 days; the 30,700+ "zombie cases" taking over a year aren't even visible here.

---

**Lessons:** While Boston's 311 system resolves most requests quickly, the data reveals important questions about equity and efficiency. The 30,700+ "zombie cases" and 3,114 impossible records suggest that aggregate statistics can mask service delivery challenges worth investigating further.

\newpage

## Appendix A: Data Dictionary

### Original Variables Used

| Variable | Description |
|:---------|:------------|
| `OPEN_DT` | Date and time the request was opened |
| `CLOSED_DT` | Date and time the request was closed |
| `REASON` | Broad category of the service request |
| `TYPE` | Specific type of service request |
| `BRA_PD` | Boston neighborhood (Planning District) |
| `SOURCE` | How the request was submitted |

### New Variables Created

| Variable | Description | How Created |
|:---------|:------------|:------------|
| `response_time_days` | Days between open and close | `as.numeric(CLOSED_DT - OPEN_DT)` |
| `day_of_week` | Day of week request was opened | `wday(OPEN_DT, label = TRUE)` |
| `case_status` | Data quality flag | `case_when()` |

\newpage

## Appendix B: Annotated R Syntax

```{r show-code, echo=TRUE, eval=FALSE}
# SETUP
library(tidyverse); library(lubridate); library(scales)
boston_311 <- read_csv("boston_311.csv.csv")

# CREATE NEW VARIABLES
boston_311 <- boston_311 %>%
  mutate(
    response_time_days = as.numeric(CLOSED_DT - OPEN_DT),
    day_of_week = wday(OPEN_DT, label = TRUE, abbr = FALSE),
    case_status = case_when(
      response_time_days < 0 ~ "Impossible",
      response_time_days > 365 ~ "Zombie",
      is.na(response_time_days) ~ "Still Open",
      TRUE ~ "Valid"))

# FUN FACTS 1-2: Response time stats & top neighborhoods
median(boston_311$response_time_days, na.rm = TRUE)
mean(boston_311$response_time_days, na.rm = TRUE)
boston_311 %>% filter(!is.na(BRA_PD)) %>% count(BRA_PD, sort = TRUE) %>% head(5)

# FUN FACTS 3-4: Categories & day of week
n_distinct(boston_311$REASON)
boston_311 %>% count(day_of_week) %>% mutate(pct = n / sum(n) * 100)

# FUN FACTS 5-7: Max response, impossible cases, zombie cases
max(boston_311$response_time_days, na.rm = TRUE)
boston_311 %>% filter(response_time_days < 0) %>% nrow()
boston_311 %>% filter(response_time_days > 365) %>% nrow()

# FUN FACT 8: Top request sources
table(boston_311$SOURCE)

# FIGURE 1: Mean Response Time by Neighborhood
top_neighborhoods <- boston_311 %>%
  filter(!is.na(BRA_PD)) %>% 
  count(BRA_PD, sort = TRUE) %>% 
  head(10) %>% 
  pull(BRA_PD)

neighborhood_stats <- boston_311 %>%
  filter(BRA_PD %in% top_neighborhoods, 
         response_time_days >= 0, 
         response_time_days <= 365) %>%
  group_by(BRA_PD) %>%
  summarise(
    mean_response = mean(response_time_days, na.rm = TRUE),
    request_count = n()
  )

neighborhood_stats %>%
  ggplot(aes(x = reorder(BRA_PD, mean_response), y = mean_response)) +
  geom_col(aes(fill = mean_response), color = "white", width = 0.7) +
  geom_text(aes(label = paste0(round(mean_response, 1), " days")), 
            hjust = -0.1, size = 2.5, fontface = "bold") +
  geom_text(aes(label = paste0("n=", format(request_count, big.mark = ","))), 
            y = 0.02, hjust = 0, size = 2, color = "white") +
  coord_flip() +
  scale_fill_gradient(low = "#9B59B6", high = "#E74C3C", guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.18))) +
  labs(title = "Average Response Times Vary Across Neighborhoods",
       subtitle = "Top 10 neighborhoods by volume",
       x = NULL, y = "Mean Response Time (Days)") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 8, color = "gray40"),
        axis.text.y = element_text(size = 8, face = "bold"),
        panel.grid.major.y = element_blank())

# FIGURE 2: Response Time Distribution
valid_responses <- boston_311$response_time_days[
  boston_311$response_time_days >= 0 & 
  boston_311$response_time_days <= 30]

median_val <- median(valid_responses, na.rm = TRUE)
mean_val <- mean(valid_responses, na.rm = TRUE)

boston_311 %>%
  filter(response_time_days >= 0, response_time_days <= 30) %>%
  ggplot(aes(x = response_time_days)) +
  geom_histogram(binwidth = 1, fill = "#9B59B6", color = "white") +
  geom_vline(xintercept = median_val, linetype = "dashed", 
             color = "#E74C3C", linewidth = 1) +
  geom_vline(xintercept = mean_val, linetype = "dotted", 
             color = "#3498DB", linewidth = 1) +
  annotate("label", x = 15, y = 500000, 
           label = paste0("Median = ", round(median_val, 1), " day"), 
           fill = "#E74C3C", color = "white", 
           fontface = "bold", size = 2.5, label.size = 0) +
  annotate("label", x = 15, y = 380000, 
           label = paste0("Mean = ", round(mean_val, 1), " days"), 
           fill = "#3498DB", color = "white", 
           fontface = "bold", size = 2.5, label.size = 0) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Most Requests Resolved Within One Week",
       subtitle = "Distribution of response times (0-30 days)",
       x = "Response Time (Days)", y = "Number of Requests") +
  theme_minimal(base_size = 9) +
  theme(plot.title = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 8, color = "gray40"))
```
